---
title: "RNAseq_Analysis"
author: "Tzu L. Phang"
date: "April 29, 2015"
output: 
  pdf_document:
    latex_engine: lualatex
    toc: true
    toc_depth: 2
    number_sections: true
    fig_width: 5
    fig_height: 4
    fig_caption: true
    highlight: tango
---

RNAseq Analysis
===============

***
Background
----------


This tutorial outline an RNAseq analysis routine conducted in my R + Biocondcutor data analysis course (BIOS6660) using one of my collaborator's data set, Dr. Eric Schmidt.  Since RNAseq dataset is typically very large, we extracted only Chromosome 19 to ease perform the analysis on local computer (mostly laptops)

## Experimental Design

Read in the descriptive data file and run summary statistics

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
## Read in descriptive data file
targets = read.table('./data/targets.txt', header = T)
## printing file table
knitr::kable(targets)
```

It seems like we have a total of 6 FASTQ files and 2 experimental grouping (CLP vs. Sham) in mouse sample


Mapping FASTQ to genome
-----------------------

We will be using QuasR, which is an extremely versatile NGS mapping and postprocessing pipeline for RNA-seq It uses Rbowtie for upgapped alignments adn SpliceMap for spliced alignments

Note: QuasR is trying to be clever: if it finds BAM files already exists then it will not generate the file again.  Therefore, to do a fresh run, need to delete everything in the "result" folder ...

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE,eval=FALSE}
library(QuasR)
targets = read.table("./data/targets.txt", header = T)
write.table(targets[,1:2], 'data/QuasR_samples.txt', row.names=F, quote=F, sep='\t')
sampleFile = "./data/QuasR_samples.txt"
genomeFile = "./data/Mouse.chromosome.19.fa"
results = "./result"
cl = makeCluster(10)


## Single command to index reference, align all samples and generate BAM files
proj <- qAlign(sampleFile, genome=genomeFile, maxHits=1, splicedAlignment=T, alignmentsDir=results, clObj=cl, cacheDir=results)
# Note: splicedAlignment should be set to TRUE when the reads are >=50nt long  
alignstats <- alignmentStats(proj) # Alignment summary report
#knitr::kable(alignstats)
```

***
## Alignment Summary

The following enumerates the number of reads in each FASTQ file and how many of them aligned to the reference. Note: the percentage of aligned reads is 100% in this particular example because only alignable reads were selected when generating the sample FASTQ files for this exercise. For QuasR this step can be omitted because the qAlign function generats this information automatically.
 
```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
library(ShortRead); library(Rsamtools)
## Extract bam file names:
bam.filenames = proj@alignments$FileName

Nreads <- countLines(dirPath="./data/", pattern=".fastq$")/4
bfl <- BamFileList(bam.filenames, yieldSize=50000, index=character())
Nalign <- countBam(bfl, param=ScanBamParam(flag=scanBamFlag(isUnmappedQuery=F)))
(read_statsDF <- data.frame(FileName=names(Nreads), Nreads=Nreads, Nalign=Nalign$records, Perc_Aligned=Nalign$records/Nreads*100))
write.table(read_statsDF, "./result/read_statsDF.txt", row.names=FALSE, quote=FALSE, sep="\t")

knitr::kable(read_statsDF)

```
 
***
## Quality Report

The following shows how to create read quality reports with QuasR’s qQCReport function or with the custom seeFastq function


```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE, eval=FALSE}
qQCReport(proj, pdfFilename="qc_report.pdf")
source("http://faculty.ucr.edu/~tgirke/Documents/R_BioCond/My_R_Scripts/fastqQuality.R")
myfiles <- paste0("data/", targets$FileName); names(myfiles) <- targets$SampleName
fqlist <- seeFastq(fastq=myfiles, batchsize=50000, klength=8)
pdf("RNAseq_Results/fastqReport.pdf", height=18, width=4*length(myfiles)); seeFastqPlot(fqlist); dev.off()

```

***
## Store Annotations in TranscriptDb

Storing annotation ranges in TranscriptDb databases makes many operations more robust and convenient

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}

library(GenomicFeatures)
txdb <- makeTranscriptDbFromGFF(file="./data/genes.gtf",
                                format="gtf",
                                dataSource="mm9",
                                species="Mus musculus")
saveDb(txdb, file="./data/mouse_annotation.sqlite")
txdb <- loadDb("./data/mouse_annotation.sqlite") 
eByg <- exonsBy(txdb, by="gene")
```

***
## Read Counting with countOverlaps

The gene signal was obtained by overlapping sequened reads onto the pre-defined gene region ranges.  The resulting count table was further filtered to remove genes with zero counts.

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
samples <- as.character(targets$Factor_long)
samplespath <- bam.filenames
names(samplespath) <- samples
countDF <- data.frame(row.names=names(eByg))
for(i in samplespath) {
  aligns <- readGAlignmentsFromBam(i) # Substitute next two lines with this one.
  seqnames(aligns) = rep('chr19', length(aligns))
  counts <- countOverlaps(eByg, aligns, ignore.strand=TRUE)
  countDF <- cbind(countDF, counts)
}
colnames(countDF) <- samples

## Remove row with all zeros
row.sum = rowSums(countDF)
chr19.countDF = countDF[row.sum != 0,]

chr19.countDF[1:4,]
write.table(chr19.countDF, "./result/chr19_countDF", quote=FALSE, sep="\t", col.names = NA)
countDF <- read.table("./result/chr19_countDF")
```

***
## Simple RPKM Normalization

RPKM: reads per kilobase of exon model per million mapped reads

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
returnRPKM <- function(counts, gffsub) {
  geneLengthsInKB <- sum(width(reduce(gffsub)))/1000 # Length of exon union per gene in kbp
  millionsMapped <- sum(counts)/1e+06 # Factor for converting to million of mapped reads.
  rpm <- counts/millionsMapped # RPK: reads per kilobase of exon model.
  rpkm <- rpm/geneLengthsInKB # RPKM: reads per kilobase of exon model per million mapped reads.
  return(rpkm)
}
countDFrpkm <- apply(countDF, 2, function(x) returnRPKM(counts=x, gffsub=eByg[rownames(countDF)]))
countDFrpkm[1:4,]
```

***
## Reproducibility Check by Sample-Wise Clustering

QC check of the sample reproducibility by computing a correlating matrix and plotting it as a tree. Note: the plotMDS function from edgeR is a more robust method for this task.


```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
library(ape) 
d <- cor(countDFrpkm, method="spearman")
hc <- hclust(dist(1-d))
plot.phylo(as.phylo(hc), type="p", edge.col=4, edge.width=3, show.node.label=TRUE, no.margin=TRUE)

```

As expected, CLP and Sham samples clustered together respectively

***
## Identify DEGs with Simple Fold Change Method

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
## Compute mean values for replicates
source("http://faculty.ucr.edu/~tgirke/Documents/R_BioCond/My_R_Scripts/colAg.R")
countDFrpkm_mean <- colAg(myMA=countDFrpkm, group=c(1,1,1,2,2,2), myfct=mean)
countDFrpkm_mean[1:4,]

## Log2 fold change
countDFrpkm_mean <- cbind(countDFrpkm_mean, log2ratio=log2(countDFrpkm_mean[,2]/countDFrpkm_mean[,1]))
countDFrpkm_mean <- countDFrpkm_mean[is.finite(countDFrpkm_mean[,3]), ]
degs2fold <- countDFrpkm_mean[countDFrpkm_mean[,3] >= 1 | countDFrpkm_mean[,3] <= -1,]
degs2fold[1:4,]
write.table(degs2fold, "./result/degs2fold.xls", quote=FALSE, sep="\t", col.names = NA)
degs2fold <- read.table("./result/degs2fold.xls")
```

***
## Identify DEGs with DESeq Library

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
library(DESeq)
countDF <- read.table("./result/chr19_countDF")
conds <- targets$Factor
cds <- newCountDataSet(countDF, conds) # Creates object of class CountDataSet derived from eSet class
counts(cds)[1:4, ] # CountDataSet has similar accessor methods as eSet class.
cds <- estimateSizeFactors(cds) # Estimates library size factors from count data. Alternatively, one can provide here the true library sizes with sizeFactors(cds) <- c(..., ...)
cds <- estimateDispersions(cds) # Estimates the variance within replicates
res <- nbinomTest(cds, "CLP", "Sham") # Calls DEGs with nbinomTest
res <- na.omit(res)
res2fold <- res[res$log2FoldChange >= 1 | res$log2FoldChange <= -1,]
res2foldpadj <- res2fold[res2fold$padj <= 0.05, ]
res2foldpadj[1:4,1:8]
```

***
## Identify DEGs with edgeR's Exact Method

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
library(edgeR)
countDF <- read.table("./result/chr19_countDF")
y <- DGEList(counts=countDF, group=conds) # Constructs DGEList object
y <- estimateCommonDisp(y) # Estimates common dispersion
y <- estimateTagwiseDisp(y) # Estimates tagwise dispersion
et <- exactTest(y, pair=c("CLP", "Sham")) # Computes exact test for the negative binomial distribution.
topTags(et, n=4)
edge <- as.data.frame(topTags(et, n=50000)) 
edge2fold <- edge[edge$logFC >= 1 | edge$logFC <= -1,]
edge2foldpadj <- edge2fold[edge2fold$FDR <= 0.01, ]
```

***
## Idenfity DEGs with edgeR's GLM Approach

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
library(edgeR)
countDF <- read.table("./result/chr19_countDF")
y <- DGEList(counts=countDF, group=conds) # Constructs DGEList object
## Filtering and normalization
keep <- rowSums(cpm(y)>1) >= 2; y <- y[keep, ]
y <- calcNormFactors(y)
design <- model.matrix(~0+group, data=y$samples); colnames(design) <- levels(y$samples$group) # Design matrix
## Estimate dispersion
y <- estimateGLMCommonDisp(y, design, verbose=TRUE) # Estimates common dispersions
y <- estimateGLMTrendedDisp(y, design) # Estimates trended dispersions
y <- estimateGLMTagwiseDisp(y, design) # Estimates tagwise dispersions 
## Fit the negative binomial GLM for each tag
fit <- glmFit(y, design) # Returns an object of class DGEGLM
contrasts <- makeContrasts(contrasts="CLP-Sham", levels=design) # Contrast matrix is optional
lrt <- glmLRT(fit, contrast=contrasts[,1]) # Takes DGEGLM object and carries out the likelihood ratio test. 
edgeglm <- as.data.frame(topTags(lrt, n=length(rownames(y))))
## Filter on fold change and FDR
edgeglm2fold <- edgeglm[edgeglm$logFC >= 1 | edgeglm$logFC <= -1,]
edgeglm2foldpadj <- edgeglm2fold[edgeglm2fold$FDR <= 0.01, ]
```

***
## Comparison Among DEG Results

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
source("http://faculty.ucr.edu/~tgirke/Documents/R_BioCond/My_R_Scripts/overLapper.R")
setlist <- list(edgeRexact=rownames(edge2foldpadj), edgeRglm=rownames(edgeglm2foldpadj), DESeq=as.character(res2foldpadj[,1]), RPKM=rownames(degs2fold))
OLlist <- overLapper(setlist=setlist, sep="_", type="vennsets")
counts <- sapply(OLlist$Venn_List, length)
vennPlot(counts=counts, mymain="DEG Comparison")
```

Number of common genes among all 4 methods: `r length(OLlist$Venn_List$edgeRexact_edgeRglm_DESeq_RPKM)`

***
## Heatmap of Top Ranking DEGs

```{r, cache=TRUE, results='markup', message=FALSE, warning=FALSE}
library(lattice); library(gplots)
y <- countDFrpkm[rownames(edgeglm2foldpadj)[1:15],]
colnames(y) <- targets$Factor
y <- t(scale(t(as.matrix(y))))
y <- y[order(y[,1]),]
levelplot(t(y), height=0.2, col.regions=colorpanel(40, "darkblue", "yellow", "white"), main="Expression Values (DEG Filter: FDR 1%, FC > 2)", colorkey=list(space="top"), xlab="", ylab="Gene ID")


```